---
title: "R Refresher Notebook"
author: "Jay Narhan"
output: html_notebook
---

## Creating Atomic Vectors
Vectors are a central component of R, not just another data structure. A vector can contain either numbers, strings, or logical values but not a mixture.
The c(...) operator can construct a vector from simple elements:

```{r}
dbl_vars <- c(1, 2.5, 4.7)   # Defining doubles (aka numeric)
int_vars <- c(1L, 6L, 10L)   # 'L' denotes integer
log_vars <- c(T, F, FALSE)
char_vars <- c("these are", "some strings")
```

c(...) will flatten vector if arguments are themselves vectors or nested vectors:

```{r}
concat_vars <- c(c(1,2,3), c(4,5,6), c(7,8, c(9,10)))
concat_vars
```

### Coercion

Data elements can coexist in a *vector* only if they have the same "mode" (same atomic data type):
```{r}
print(int_vars)
is.atomic(int_vars)
mode(int_vars)              # Notice that integers and doubles are described as "numeric""

print(c(1, F, 'test'))      # Downgraded to characters in order to accomdate all constituent elements
is.atomic(c(1, F, 'test'))  # Note: vector is downgraded to atomic values, in this case character data
mode(c(1, F, 'test'))
```

Technically, above is an example of coercion of data types and will often happen automatically. It can be quite helpful, for example finding the total number of Trues in vector via ```sum(c(TRUE, TRUE, FALSE, FALSE, FALSE))``` which yields `r sum(c(TRUE, TRUE, FALSE, FALSE, FALSE))`, or to find the proportion of Trues by using ```mean(c(TRUE, TRUE, FALSE, FALSE, FALSE))``` which is `r mean(c(TRUE, TRUE, FALSE, FALSE, FALSE))`. In both cases, logical values have been coerced to numeric values (1's and 0's).

In programming, if any confusion could arise, you should explicitly cast the data type using e.g. as.integer()

```{r}
as.integer(c("-1", 1))
```

### Scalars
```{r}
is.vector(10)               # In R - no scalars (0d), 'scalars' stored as 1d elements i.e. atomic vector
str(10)
str(10L)
```

### Tests and Types

```{r}
typeof(int_vars)
mode(int_vars)       # notice the mode for doubles or integers is numeric
is.integer(int_vars)

typeof(dbl_vars)
mode(dbl_vars)       # notice the mode for doubles or integers is numeric
is.integer(dbl_vars)
is.double(dbl_vars)
```
## Lists

Elements within a list can be of any type. Create using list() instead of c().

```{r}
x <- list( 1:5, 'a', c(T, FALSE, F, TRUE, T), c(2.3, 7.9))
str(x)
```

Lists can contain lists
```{r}
recu_list <- list( list( list( c(1,2,3), 'a')))
str(recu_list)

str(recu_list)
is.recursive(recu_list)
```

```{r}
x <- list( list(1,2), c(3,4))
str(x)
```

If combining atomic vectors and lists, c() will coerce the vectors to lists, before combining them:
```{r}
y <- c(list(1,2), c(3,4))
str(y)

# Explaination: c(3,4) is coerced to being a list (a list of 2 items with numeric values (scalars) 3 and 4 for each list entry)
str(as.list(c(3,4)))

# These two lists are then flattened via c() into one list. Remember c() flattens stuff.
```

unlist() can be used to convert a list into an atomic vector (if elements in the list are different types, coercion rules are applied).

```{r}
str(y)
unlist(y)

str(recu_list)
unlist(recu_list)
```

## Attributes

### Factors
```{r}
z <- read.csv(text = "value\n12\n1\n. \n9")
z
```

```{r}
str(z)
```

The '. ' value, makes R read it as a character vector, which is loaded into R by the default action of treating strings as factors.

```{r}
z <- read.csv(text = "value\n12\n1\n. \n9", na.strings = '. ')
z
```

```{r}
str(z)
```

The default action of treating strings as factors is sub-optimal because the loading functions may not be aware of all the possible factors. Best to load with `StringsAsFactors = False` (avoid changing the global option as this may have ramifications when integrating in external code).

## Matricies & Arrays

Adding a dim() attribute to atomic vector will create a multi-d array

```{r}
c <- 1:6
str(c)

dim(c) <- c(2,3)
str(c)
c
```

Generally use matrix() and array()

```{r}
d <- matrix(1:6, ncol = 3, nrow = 2)
d

dim(d) <- c(3,2)
d
```

```{r}
length(d)
dim(d)
nrow(d)
ncol(d)
```

```{r}
rownames(d) <- c("A", "B", "C")
colnames(d) <- c("a", "b")
d
```

```{r}
e <- array(1:6, dim = c(2,3,2))
e
```

```{r}
length(e)
dim(e)
```


## Dataframes

A list of equal-length vectors.

```{r}
df <- data.frame(x = 1:3, y = c('a','b','c'))
str(df)
```

The default behaviour is to turn strings into factors .. counter by:

```{r}
df <- data.frame(x = 1:3, y = c('a','b','c'), stringsAsFactors = FALSE)
str(df)
df
```

### Combining Data Frames

```{r}
z <- data.frame(z = 3:1)
z
```


```{r}
cbind(df,z)
```


```{r}
rbind(df, data.frame(x = 10:13, y = c('d','e','f', 'g')))

str(rbind(df, data.frame(x = 10:13, y = c('d','e','f', 'g'))))
```